---
title: "Determine number of clusters"
author: "Stefano De Sabbata"
date: "`r lubridate::now()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

```{r libraries, warning=FALSE, message=FALSE}
library(tidyverse)
library(magrittr)
library(knitr)
library(GGally)
```



## Data

Load the [selected census variables](50-prep/111-classification-variable-selection.html).

```{r}
census_data_trans_selected <-
  read_csv(
    "../storage/dati-cpa_2011_all-trans-selected-v0_0_4.csv",
    col_types = paste(c(rep("c", 12), rep("d", 116)), collapse="")
  )
```

## Determine number of clusters

```{r}
# Data for elbow method
data_for_testing <-
  census_data_trans_selected %>%
  dplyr::select(P1_norm_log10_std:E30_E31_norm_std) %>% 
  slice_sample(prop = 0.01)
```

```{r}
# Calculate WCSS and silhouette
# for k = 2 to 50
# Set up two vectors where to store
# the calculated WCSS and silhouette value
testing_wcss <- rep(NA, 50)
testing_silhouette <- rep(NA, 50)

# for k = 2 to 50
for (testing_k in 2:50){
  #cat(testing_k, "\n")
  
  # Calculate kmeans
  kmeans_result <- 
    stats::kmeans(data_for_testing, centers = testing_k, iter.max = 50)
  
  # Extract WCSS
  # and save it in the vector
  testing_wcss[testing_k] <- kmeans_result %$% tot.withinss
  
  # Calculate average silhouette
  # and save it in the vector
  testing_silhouette[testing_k] <- 
    kmeans_result %$% cluster %>%
    cluster::silhouette(
      data_for_testing %>% dist()
    ) %>%
    magrittr::extract(, 3) %>% mean()
}
```

```{r, eval=FALSE}
# Calculate the gap statistic using bootstrapping
testing_gap <- 
  cluster::clusGap(data_for_testing, FUN = kmeans, 
    K.max = 50, B = 50
  )
```

### Plots

```{r}
plot(2:50, testing_wcss[2:50], type="b", xlab="Number of Clusters", 
     ylab="WCSS", xlim=c(1,50)) +
abline(v = 8, col = "red") +
abline(v = 20, col = "red")
```

```{r}
plot(2:50, testing_silhouette[2:50], type="b", xlab="Number of Clusters", 
     ylab="Silhouette", xlim=c(1,50)) +
abline(v = 8, col = "red") +
abline(v = 20, col = "red")
```

```{r, eval=FALSE}
plot(2:50, testing_gap[["Tab"]][2:50, "gap"], type="b", xlab="Number of Clusters", 
     ylab="Gap", xlim=c(1,50)) +
abline(v = 8, col = "red") +
abline(v = 20, col = "red")
```

## Conclusions

This is a **draft**, the analysis is still on-going.

The analysis above suggests that 8 and 20 might be reasonably good as number of clusters to create the geodemographic classification.



## Acknowledgements

This analysis uses data from [ISTAT](https://www.istat.it/it/archivio/104317) distributed under [CC BY 3.0 IT](https://creativecommons.org/licenses/by/3.0/it/deed.en) (see also [legal notice](https://www.istat.it/it/note-legali)).



# Session info

```{r}
sessionInfo()
```

