---
title: "Clustering"
author: "Stefano De Sabbata"
date: "`r lubridate::now()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

```{r libraries, warning=FALSE, message=FALSE}
library(tidyverse)
library(magrittr)
library(knitr)
library(sf)
library(patchwork)
library(ggrepel)
```



## Data

Load the [selected census variables](100-prep/111-classification-variable-selection.html).

```{r}
census_data_trans_selected <-
  read_csv(
    "../storage/dati-cpa_2011_all-trans-selected-v0_0_4.csv",
    col_types = paste(c(rep("c", 12), rep("d", 116)), collapse="")
  )
```


## Sensitivity analysis

Assess WCSS and BCSS as each variable is iteratively removed. Where the removal of a variable leads lower WCSS and higher BCSS than average, that indicates that removing that variable from the clustering process would lead to more homogeneousand distinct clusters.

```{r}
ieac_k08_v0_0_4_sensitivity_analysis <-
  readRDS("../storage/ieac_k08_v0_0_4_sensitivity_analysis.rds")
```


### Plot

```{r, fig.height=7, fig.width=7}
sensitivity_plot_wcss <-
  ieac_k08_v0_0_4_sensitivity_analysis %>% 
  ggplot(aes(
    x = wcss,
    y = census_variable, 
    group=1
  )) +
  geom_point(stat='summary', fun.x=sum) +
  stat_summary(fun.x=sum, geom="line") +
  theme_bw()

sensitivity_plot_bcss <-
  ieac_k08_v0_0_4_sensitivity_analysis %>% 
  ggplot(aes(
    x = bcss,
    y = census_variable, 
    group=1
  )) +
  geom_point(stat='summary', fun.x=sum) +
  stat_summary(fun.x=sum, geom="line") +
  theme_bw()

sensitivity_plot_wcss + sensitivity_plot_bcss
```

Highlight variables in the top-left corner, as where the removal of a variable leads lower WCSS and higher BCSS than average, that indicates that removing that variable from the clustering process would lead to more homogeneous and distinct clusters.

```{r, fig.height=7, fig.width=7}
ieac_k08_v0_0_4_sensitivity_analysis %>%
  mutate(
    plot_label = if_else(
      ((bcss) > ((0.75 * wcss) - 192500)),
      census_variable, ""
    )
  ) %>% 
  ggplot(aes(
    x = wcss,
    y = bcss,
    label = plot_label
  )) +
  geom_point() +  
  geom_text_repel(
    colour = "#999999", 
    size = 3,
    force = 1,
    force_pull = 0.1,
    box.padding = 0.5,
    min.segment.length = 0,
    segment.curvature = -0.1
  ) + 
  theme_bw()
```



### Variables to exclude

| Variable code | Variable description                                                                                                                                                      | Normalisation code | Normalisation description                                                                                                                                        |
|---------------|---------------------------------------------------------------------------------------------------------------------------------------------------------------------------|--------------------|------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| P8            | Popolazione residente - divorziati/e                                                                                                                                      | P1                 | Popolazione residente - totale                                                                                                                                   |
| P30           | Popolazione residente - maschi - età < 5 anni                                                                                                                             | P14                | Popolazione residente - età < 5 anni                                                                                                                             |
| P36           | Popolazione residente - maschi - età 30 - 34 anni                                                                                                                         | P20                | Popolazione residente - età 30 - 34 anni                                                                                                                         |
| P66           | Popolazione residente - maschi di 15 anni e più disoccupata in cerca   nuova occupazione                                                                                  | P62                | Popolazione residente - totale di 15 anni e più disoccupata in cerca   nuova occupazione                                                                         |
| ST7           | Stranieri e apolidi residenti in Italia - maschi - età 30 - 54 anni                                                                                                       | ST4                | Stranieri e apolidi residenti in Italia - età 30 - 54 anni                                                                                                       |



## Clustering

```{r}
census_data_kmeans_k08 <- 
  census_data_trans_selected %>%
  select(P1_norm_log10_std:E30_E31_norm_std) %>%
  select(
    -P8_norm_log10_std, -P30_norm_std, 
    -P36_norm_std, -P66_norm_std,
    -ST7_norm_std
  ) %>% 
  kmeans(
    centers = 8, iter.max = 5000,
    algorithm = "Lloyd"
  )
```

```{r}
census_data_ieac_k08 <- 
  census_data_trans_selected %>%
  tibble::add_column(
    ieac_k08 = census_data_kmeans_k08 %$% cluster %>% as.character()
  )
```

## Save values

```{r}
census_data_kmeans_k08 %>% 
  saveRDS("../storage/ieac_k08_kmeans_object-v0_0_4_6.rds")
```


```{r}
census_data_ieac_k08 <-
  census_data_ieac_k08 %>% 
  left_join(
    read_csv(
      "../storage/dati-cpa_2011_all.csv",
      col_types = paste(c(rep("c", 12), rep("i", 140)), collapse="")
    ) %>% 
    select(SEZ2011, P1)
  ) %>% 
  mutate(ieac_k08 = if_else(P1 == 0 | P1 >= 10, ieac_k08, NA_character_))

# Check
census_data_ieac_k08 %>% 
  filter(P1 > 0) %>% 
  pull(P1) %>% 
  min()

census_data_ieac_k08 <-
  census_data_ieac_k08 %>% 
  select(-P1)
```

```{r}
census_data_ieac_k08 %>% 
  write_csv("../storage/ieac_k08-with-vars-v0_0_4_6.csv") 

census_data_ieac_k08 %>% 
  select(CODREG:CODASC, ieac_k08) %>%
  write_csv("../storage/ieac_k08-no-vars-v0_0_4_6.csv") 
```

```{r}
census_data_ieac_k08_geo <-
  readRDS( "../storage/italian-enumeration-areas-2011.rds") %>%
  left_join(
    census_data_ieac_k08 %>% 
    select(CODREG:CODASC, ieac_k08)
  )

census_data_ieac_k08_geo %>% 
  saveRDS("../storage/ieac_k08-with-vars-v0_0_4_6.rds")

census_data_ieac_k08_geo %>% 
  write_sf("../storage/ieac_k08-with-vars-v0_0_4_6.geojson")

dir.create("../storage/ieac_k08-with-vars-v0_0_4_6")
census_data_ieac_k08_geo %>% 
  write_sf("../storage/ieac_k08-with-vars-v0_0_4_6/ieac_k08-with-vars-v0_0_4_6.shp")
```



## Conclusions

This is a **draft**, the analysis is still on-going.



## Acknowledgements

This analysis uses data from [ISTAT](https://www.istat.it/it/archivio/104317) distributed under [CC BY 3.0 IT](https://creativecommons.org/licenses/by/3.0/it/deed.en) (see also [legal notice](https://www.istat.it/it/note-legali)).



# Session info

```{r}
sessionInfo()
```

