---
title: "Clustering"
author: "Stefano De Sabbata"
date: "`r lubridate::now()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

```{r libraries, warning=FALSE, message=FALSE}
library(tidyverse)
library(magrittr)
library(knitr)
library(sf)
library(patchwork)
library(ggrepel)
library(e1071)
```



## Data

Load the [selected census variables](100-prep/111-classification-variable-selection.html).

```{r}
census_data_trans_selected <-
  read_csv(
    "../storage/dati-cpa_2011_all-trans-selected-v0_0_4.csv",
    col_types = paste(c(rep("c", 12), rep("d", 116)), collapse="")
  )
```


## Sensitivity analysis

Assess WCSS and BCSS as each variable is iteratively removed. Where the removal of a variable leads lower WCSS and higher BCSS than average, that indicates that removing that variable from the clustering process would lead to more homogeneousand distinct clusters.

```{r}
ieac_k20_v0_0_4_sensitivity_analysis <-
  readRDS("../storage/ieac_k20_v0_0_4_sensitivity_analysis.rds")
```


### Plot

```{r, fig.height=7, fig.width=7}
sensitivity_plot_wcss <-
  ieac_k20_v0_0_4_sensitivity_analysis %>% 
  ggplot(aes(
    x = wcss,
    y = census_variable, 
    group=1
  )) +
  geom_point(stat='summary', fun.x=sum) +
  stat_summary(fun.x=sum, geom="line") +
  theme_bw()

sensitivity_plot_bcss <-
  ieac_k20_v0_0_4_sensitivity_analysis %>% 
  ggplot(aes(
    x = bcss,
    y = census_variable, 
    group=1
  )) +
  geom_point(stat='summary', fun.x=sum) +
  stat_summary(fun.x=sum, geom="line") +
  theme_bw()

sensitivity_plot_wcss + sensitivity_plot_bcss
```

Highlight variables in the top-left corner, as where the removal of a variable leads lower WCSS and higher BCSS than average, that indicates that removing that variable from the clustering process would lead to more homogeneous and distinct clusters.

```{r, fig.height=7, fig.width=7}
ieac_k20_v0_0_4_sensitivity_analysis %>%
  mutate(
    plot_label = if_else(
      ((bcss) > ((1.0 * wcss) - 234500)) | (bcss > 93500),
      census_variable, ""
    )
  ) %>% 
  ggplot(aes(
    x = wcss,
    y = bcss,
    label = plot_label
  )) +
  geom_text_repel(
    colour = "#999999", 
    size = 3,
    force = 1,
    force_pull = 0.1,
    box.padding = 0.5,
    min.segment.length = 0,
    segment.curvature = -0.1
  ) + 
  geom_point() +  
  theme_bw()
```



### Variables to reconsider

| Variable code | Variable description                                                                                                                                                      | Normalisation code | Normalisation description                                                                                                                                        |
|---------------|---------------------------------------------------------------------------------------------------------------------------------------------------------------------------|--------------------|------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| P30           | Popolazione residente - maschi - età < 5 anni                                                                                                                             | P14                | Popolazione residente - età < 5 anni                                                                                                                             |
| P31           | Popolazione residente - maschi - età 5 - 9 anni                                                                                                                           | P15                | Popolazione residente - età 5 - 9 anni                                                                                                                           |
| P36           | Popolazione residente - maschi - età 30 - 34 anni                                                                                                                         | P20                | Popolazione residente - età 30 - 34 anni                                                                                                                         |
| P38           | Popolazione residente - maschi - età 40 - 44 anni                                                                                                                         | P22                | Popolazione residente - età 40 - 44 anni                                                                                                                         |
| P66           | Popolazione residente - maschi di 15 anni e più disoccupata in cerca   nuova occupazione                                                                                  | P62                | Popolazione residente - totale di 15 anni e più disoccupata in cerca   nuova occupazione                                                                         |
| E11           | Edifici ad uso residenziale costruiti dal 1961 al 1970                                                                                                                    | E3                 | Edifici ad uso residenziale                                                                                                                                      |
| E23           | Edifici ad uso residenziale da 3 a 4 interni                                                                                                                              | E3                 | Edifici ad uso residenziale                                                                                                                                      |

Based on the analysis above, all variables P30 to P45 will be removed, along with P46. All other variables will still be retained.

## Clustering

```{r}
census_data_kmeans_k20_RData_file <-
  "../storage/ieac_k20_kmeans_object-v0_0_4_16.RData"

if (file.exists(census_data_kmeans_k20_RData_file)) {
  load(census_data_kmeans_k20_RData_file)
} else{
  
  census_data_kmeans_k20 <- 
    census_data_trans_selected %>%
    select(P1_norm_log10_std:E30_E31_norm_std) %>%
    select(
      -P30_norm_std, -P31_norm_std, -P32_norm_std, -P33_norm_std,
      -P34_norm_std, -P35_norm_std, -P36_norm_std, -P37_norm_std,
      -P38_norm_std, -P39_norm_std, -P40_norm_std, -P41_norm_std,
      -P42_norm_std, -P43_norm_std, -P44_norm_std, -P45_norm_std,
      -P46_norm_log10_std
    ) %>% 
    # kmeans(
    #   centers = 20,
    #   iter.max = 5000
    # )
    e1071::bclust(
      centers = 20,
      base.centers = 50,
      iter.base = 100,
      iter.max = 50000,
      resample = FALSE,
      algorithm="Lloyd"
    )

  save(
    census_data_kmeans_k20, 
    file = census_data_kmeans_k20_RData_file
  )
  
}
```

```{r}
centers_hclust  <- 
  hclust(
    census_data_kmeans_k20 %$%
      centers %>% 
      dist(), 
    "ward.D2"
  )
plot(centers_hclust, hang =- 1)
```

Generate cluster names.

```{r}
#cut_tk03 <- cutree(centers_hclust, k = 3) %>% as.vector()
cut_tk06 <- cutree(centers_hclust, k = 6) %>% as.vector()
cut_tk20 <- cutree(centers_hclust, k = 20) %>% as.vector()

clust_names <- rep(NA, length(cut_tk20))

# id_at_tk03 <- 1
# for(i_tk03 in unique(cut_tk03)){
#   at_tk06 <- cut_tk06[cut_tk03 == i_tk03]
  id_at_tk06 <- 1
  #for(i_tk06 in unique(at_tk06)){
  for(i_tk06 in unique(cut_tk06)){
    at_tk20 <- cut_tk20[cut_tk06 == i_tk06]
    id_at_tk20 <- 1
    if(length(at_tk20) == 1){
      i_tk20 <- at_tk20[1]
      #clust_name <- paste0(id_at_tk03, letters[id_at_tk06])
      clust_name <- paste0(id_at_tk06)
      clust_names[i_tk20] <- clust_name
      cat(i_tk20, " - ", clust_name, "\n")
    }else{
      for(i_tk20 in at_tk20){
        #clust_name <- paste0(id_at_tk03, letters[id_at_tk06], id_at_tk20)
        clust_name <- paste0(id_at_tk06, letters[id_at_tk20])
        clust_names[i_tk20] <- clust_name
        cat(i_tk20, " - ", clust_name, "\n")
        id_at_tk20 <- id_at_tk20 + 1
      }
    }
    id_at_tk06 <- id_at_tk06 + 1
  }
#   id_at_tk03 <- id_at_tk03 + 1
# }
rm(
  at_tk20, at_tk06,
  clust_name,
  cut_tk20, cut_tk06, #cut_tk03, 
  id_at_tk20, id_at_tk06, #id_at_tk03,
  i, i_tk20, i_tk06 #, i_tk03
)
```

```{r}
census_data_ieac_k20 <- 
  census_data_trans_selected %>%
  tibble::add_column(
    ieac_k20 = census_data_kmeans_k20 %$% cluster
  ) %>% 
  mutate(
    ieac_k20h = clust_names[ieac_k20]
  ) %>% 
  mutate(
     ieac_k20 = ieac_k20 %>% as.character()
  )
```



## Save values


```{r}
census_data_ieac_k20 <-
  census_data_ieac_k20 %>% 
  left_join(
    read_csv(
      "../storage/dati-cpa_2011_all.csv",
      col_types = paste(c(rep("c", 12), rep("i", 140)), collapse="")
    ) %>% 
    select(SEZ2011, P1)
  ) %>% 
  mutate(
    ieac_k20 = if_else(P1 == 0 | P1 >= 10, ieac_k20, NA_character_),
    ieac_k20h = if_else(P1 == 0 | P1 >= 10, ieac_k20h, NA_character_)
  )

# Check
census_data_ieac_k20 %>% 
  filter(P1 > 0 & !is.na(ieac_k20)) %>% 
  pull(P1) %>% 
  min()

census_data_ieac_k20 <-
  census_data_ieac_k20 %>% 
  select(-P1)
```

```{r}
census_data_ieac_k20 %>% 
  write_csv("../storage/ieac_k20-with-vars-v0_0_4_16.csv") 

census_data_ieac_k20 %>% 
  select(CODREG:CODASC, ieac_k20, ieac_k20h) %>%
  write_csv("../storage/ieac_k20-no-vars-v0_0_4_16.csv") 
```

```{r}
census_data_ieac_k20_geo <-
  readRDS( "../storage/italian-enumeration-areas-2011.rds") %>%
  left_join(
    census_data_ieac_k20 %>% 
    select(CODREG:CODASC, ieac_k20, ieac_k20h)
  )

census_data_ieac_k20_geo %>% 
  saveRDS("../storage/ieac_k20-with-vars-v0_0_4_16.rds")

census_data_ieac_k20_geo %>% 
  write_sf("../storage/ieac_k20-with-vars-v0_0_4_16.geojson")

dir.create("../storage/ieac_k20-with-vars-v0_0_4_16")
census_data_ieac_k20_geo %>% 
  write_sf("../storage/ieac_k20-with-vars-v0_0_4_16/ieac_k20-with-vars-v0_0_4_16.shp")
```



## Conclusions

This is a **draft**, the analysis is still on-going.



## Acknowledgements

This analysis uses data from [ISTAT](https://www.istat.it/it/archivio/104317) distributed under [CC BY 3.0 IT](https://creativecommons.org/licenses/by/3.0/it/deed.en) (see also [legal notice](https://www.istat.it/it/note-legali)).



# Session info

```{r}
sessionInfo()
```

